---
title: 'fianl Project update #2'
author: "Jorge Dimas-Geniz"
date: "2022-12-01"
output: html_document
---
title: "Data analysis final project"
author: "Jorge Dimas-Geniz"
date: "2022-11-22"
output: html_document
---
```{r}
ggplot(mtss, aes(TotalStress,Hippocampus )) + geom_point()+ geom_smooth()
```
```{r}
ggplot(mtss, aes(TotalStress,Amygdala )) + geom_point()+ geom_smooth()
```
```{r}
ggplot(mtss, aes(TotalStress,Thalamus )) + geom_point()+ geom_smooth()
```
#This is based on a paper that measures the volumes of the hippocamus, amygdala, and thalamus. Stress was also measured and correlated in this study.
#Research Question: How does stress and abuse have an effect on neural development of children?
```{r}
library(ggplot2)
library(ltm)
```

```{r}
#Our dataset is from an already published paper. We are using the data to explore the correlations between children/adolescent stress level and hippocampus, amygdala, and thalamus size. 
finalProjDat<-mtss <- read_excel("C:/Users/Jorge/Downloads/mtss.xlsx")
```
#Dataset Manipulation
```{r}
#First, we created new columns for stress, amygdala volume, and total thalamus volume. This helps when setting up the graphs and statistical analyses later. 
#Stress was measured using UCLA Life Stress Interview (Hammen, 1988). Both parents and kids were interviewed and the higher of the two were included in the data with scores ranging from 0-45.
#Hippocampal and amygdala volume were measured in mm^3 x 10^3. No measurement was indicated for the thalamus, but it was assumed to be the same measurement. 
#All three brain parts were measured using structural MRI.
finalProjDat$Stress<- finalProjDat$TotalStress
finalProjDat$AmygdalaVolume<- finalProjDat$Amygdala
finalProjDat$HippocampusVolume<- finalProjDat$Hippocampus
ThalamusVolume<- function(rThal,lThal){#combining the two thalamus volumes to make a total volume
  rThal + lThal
}
finalProjDat$ThalamusVolume<- ThalamusVolume(finalProjDat$RightThalamusProper,finalProjDat$LeftThalamusProper)
```
#work on abuse on thursday
```{r}
finalProjDat$Abuse<- finalProjDat$ANY_ABUSE_CHILD_DV
class(finalProjDat$Abuse)#integer, need to be a factor in order to relevel into yes and no
finalProjDat$Abuse<- as.factor(finalProjDat$Abuse)
levels(finalProjDat$Abuse)#0 is the first level, 1 is the second
levels(finalProjDat$Abuse)<- c("no","yes")
finalProjDat$Abuse#I am unsure how to incorporate abuse into stress. We know from studies that abuse affects stress levels, but how to quantify that into the current measure of stress is where the problem lies for me. 
```

```{r}
biserial.cor(finalProjDat$Stress,finalProjDat$Abuse)#the correlation in this data between stress and abuse is -.533. It may be better off to not include the abuse variable to keep things simple
```
#Research Question 1: How does different levels of stress effect the neurological growth development of the hippocampi of children?

#Null Hypothesis 1: There will be no substantial difference in the size of hippocampi of children caused by different levels of stress.

#Alternative Hypothesis 1: Increased levels of stress in young children leads to lesser size hippocampi in children, in comparison to their peers who experience normal levels of stress.
```{r}
cor.test(finalProjDat$Stress,finalProjDat$HippocampusVolume,alternative = "less")
#this is not a significant negative correlation
ggplot(finalProjDat, aes(Stress,HippocampusVolume)) + geom_point(color = "darkmagenta") + geom_smooth(method = "lm", color = "darkslategray", se = FALSE) + theme_bw() + ylab("Hippocamus Volume") + ggtitle("Hippocampal Volume vs Stress")
```
#Analysis 1
```{r}
cor.test(finalProjDat$Stress,finalProjDat$HippocampusVolume,alternative = "less")
#Correlation of -.07
#p-value of .1843
#Nonsignificant p-value
lmHippo<- lm(HippocampusVolume~Stress*Abuse, data = finalProjDat)
summary(lmHippo)

```
# We fail to reject our null hypothesis: p-value(.1843)> (.05)

#Research question 2:  How does different levels of stress effect the neurological growth development of the amydala of children?

#Null Hypothese 2: There will be no substantial difference in the size of Amydalae of children caused by different levels of stress.

#Alternative Hypothesis 2: Increased levels of stress in young children leads to lesser size amygdalae in children, in comparison to their peers who experience normal levels of stress.
```{r}
cor.test(finalProjDat$Stress,finalProjDat$AmygdalaVolume,alternative = "less")
#this is not a significant negative correlation
ggplot(finalProjDat, aes(Stress,AmygdalaVolume)) + geom_point(color = "blue") + geom_smooth(method = "lm", color = "bisque4", se = FALSE) + theme_bw() + ylab("Amygdala Volume") + ggtitle("Whatever Title we want to use")
```
#Analysis 2
```{r}
cor.test(finalProjDat$Stress,finalProjDat$AmygdalaVolume,alternative = "less")
#Correlation of -.102
#p-value of .106
#Nonsignificant p-value
lmAmygd<- lm(AmygdalaVolume~Stress*Abuse, data = finalProjDat)
summary(lmAmygd)
```
# We fail to reject our null hypothesis: p-value(.1843)> (.05)

#Research question 3:  How does different levels of stress effect the neurological growth development of the Thalamus of children?

#Null Hypothese 2: There will be no substantial difference in the size of Thalami of children caused by different levels of stress.

#Hypothesis 3: Increased levels of stress in young children leads to lesser size thalami in children, in comparison to their peers who experience normal levels of stress.
```{r}
cor.test(finalProjDat$Stress,finalProjDat$ThalamusVolume,alternative = "less")
#this is not a significant negative correlation
ggplot(finalProjDat, aes(Stress,ThalamusVolume)) +geom_point(color = "darkorange") + geom_smooth(method = "lm", color = "darkblue", se = FALSE) + theme_bw() + ylab("Thalamus Volume") + ggtitle("Whatever title we want to use")
```
#Analysis 2
```{r}
cor.test(finalProjDat$Stress,finalProjDat$ThalamusVolume,alternative = "less")
#Correlation of -.1698
#p-value of .-07
#Nonsignificant p-value
lmThala<- lm(ThalamusVolume~Stress, data = finalProjDat)
summary(lmThala)
```
# We fail to reject our null hypothesis: p-value(-.07)> (-.05)
```{r}
#possible new statistical analysis
summary(lm(volume~stress*abuse,data = finalProjDat))
```{r}
lmHippo<- lm(HippocampusVolume~Stress*Abuse, data = finalProjDat)
summary(lmHippo)
#do this for all 3 brain parts
lmAmygd<- lm(AmygdalaVolume~Stress*Abuse, data = finalProjDat)
summary(lmAmygd)
lmThala<- lm(ThalamusVolume~Stress, data = finalProjDat)
summary(lmThala)