---
title: "Data analysis final project"
author: "Jorge Dimas-Geniz"
date: "2022-11-22"
output: html_document
---

#This is based on a paper that measures the volumes of the hippocamus, amygdala, and thalamus. Stress was also measured and correlated in this study.
#Research Question: How does stress and abuse have an effect on neural development of children?

#Our dataset is from an already published paper. We are using the data to explore the correlations between children/adolescent stress level and hippocampus, amygdala, and thalamus size. 

```{r}
finalProjDat<-read.csv("C:/Users/Chris/Downloads/PSY 4960 Programming in R/Final Project/mtss.csv")
```
#dataset manipulation.
#First, we created new columns for stress, amygdala volume, and total thalamus volume. This helps when setting up the graphs and statistical analyses later. 
#Stress was measured using UCLA Life Stress Interview (Hammen, 1988). Both parents and kids were interviewed and the higher of the two were included in the data with scores ranging from 0-45.
#Hippocampal and amygdala volume were measured in mm^3 x 10^3
#All three brain parts were measured using structural MRI.
```{r}
finalProjDat$Stress<- finalProjDat$TotalStress
finalProjDat$AmygdalaVolume<- finalProjDat$Amygdala
finalProjDat$HippocampusVolume<- finalProjDat$Hippocampus
ThalamusVolume<- function(rThal,lThal){#combining the two thalamus volumes to make a total volume
  rThal + lThal
}
finalProjDat$ThalamusVolume<- ThalamusVolume(finalProjDat$RightThalamusProper,finalProjDat$LeftThalamusProper)
finalProjDat$Abuse<- finalProjDat$ANY_ABUSE_CHILD_DV
class(finalProjDat$Abuse)#integer, need to be a factor in order to relevel into yes and no
finalProjDat$Abuse<- as.factor(finalProjDat$Abuse)
levels(finalProjDat$Abuse)#0 is the first level, 1 is the second
levels(finalProjDat$Abuse)<- c("no","yes")
finalProjDat$Abuse#I am unsure how to incorporate abuse into stress. We know from studies that abuse affects stress levels, but how to quantify that into the current measure of stress is where the problem lies for me. 
```

```{r}
#here are the graphs
library(ggplot2)
ggplot(finalProjDat, aes(Stress,ThalamusVolume)) +geom_point(color = "darkorange") + geom_smooth(method = "lm", color = "darkblue", se = FALSE) + theme_bw() + ylab("Thalamus Volume")
ggplot(finalProjDat, aes(Stress,AmygdalaVolume)) + geom_point(color = "blue") + geom_smooth(method = "lm", color = "bisque4", se = FALSE) + theme_bw() + ylab("Amygdala Volume")
ggplot(finalProjDat, aes(Stress,HippocampusVolume)) + geom_point(color = "darkmagenta") + geom_smooth(method = "lm", color = "darkslategray", se = FALSE) + theme_bw + ylab("Hippocamus Volume")
```

```{r}
library(ltm)
biserial.cor(finalProjDat$Stress,finalProjDat$Abuse)#the correlation in this data between stress and abuse is -.533. It may be better off to not include the abuse variable to keep things simple
```

```{r}
#now to do the analyses
cor.test(finalProjDat$Stress,finalProjDat$ThalamusVolume,alternative = "less")
#Pearson's product-moment correlation

#data:  finalProjDat$Stress and finalProjDat$ThalamusVolume
#t = -0.95809, df = 147, p-value = 0.1698
#alternative hypothesis: true correlation is less than 0
#95 percent confidence interval:
# -1.00000000  0.05712717
#sample estimates:
#        cor 
#-0.07877605 

#this is not a significant negative correlation
```
```{r}
cor.test(finalProjDat$Stress,finalProjDat$AmygdalaVolume,alternative = "less")
#Pearson's product-moment correlation

#data:  finalProjDat$Stress and finalProjDat$AmygdalaVolume
#t = -1.2538, df = 147, p-value = 0.106
#alternative hypothesis: true correlation is less than 0
#95 percent confidence interval:
# -1.00000000  0.03288975
#sample estimates:
#       cor 
#-0.1028623 

#this is not a significant negative correlation
```
```{r}
cor.test(finalProjDat$Stress,finalProjDat$HippocampusVolume,alternative = "less")
#Pearson's product-moment correlation

#data:  finalProjDat$Stress and finalProjDat$HippocampusVolume
#t = -0.90187, df = 147, p-value = 0.1843
#alternative hypothesis: true correlation is less than 0
#95 percent confidence interval:
# -1.00000000  0.06173368
#sample estimates:
#        cor 
#-0.07418026 

#this is not a significant negative correlation
```
